$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/paper.tex
updating syntax

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}

In the \picalc{} both scope restriction and input introduce abstractions.
The bodies of these constructs need a way of referring to their argument.
Using names as variable references is a popular option amongst humans.
However, names are cumbersome to mechanize: inserting a new variable into a context means proving that the name of such variable is different to all other variable names in context.
Moreover, names are of no significance to machines whatsoever.

The idea de Bruijn had \cite{} was to use a natural number $n$ (aka \emph{index}) to refer to the variable introduced $n$ binders ago --- and to have binders no longer introduce names.
\begin{example}
\hfill{}\\
The \picalc{} term \hfill{} $(\textbf{new} \; x) (x ( y ) . y \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new} \; y) (x \langle y \rangle . y ( z ) .\mathbf{0}))$ \\ is represented as \hfill{} $(\textbf{new}) (0 () . 0 \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new}) (1 \langle 0 \rangle . 0 () . \mathbf{0}))$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.
Humans may find this confusing.
There is however no reason not to annotate these indices with the original names bestowed upon them by humans.
Machines can then manipulate references mechanically whilst using names to present terms to humans.

A variable occurring under $n$ abstractions has $n$ possible variables to refer to.
References outside of that range are meaningless.
It is useful to rule out these nonsensical terms syntactically.
In Figure \ref{fig:var} we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
As shown in Figure \ref{fig:syntax}, we increase the \emph{depth} counter every time we create a new channel or receive some input.

@@

\begin{equation*}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar using de Bruijn indices}
\label{fig:syntax}
\end{figure}

Using type-level de Bruijn indices makes our syntax well-scoped by construction --- the semantics of our language can now be defined on the totality of the syntax.
User-friendliness can still be recovered through a function that converts processes with named variables into (possibly) processes with indexed variables.
This function recursively traverses processes, keeping track of what name is associated with each index.
If the process is ill-scoped, the function returns nothing.
To present things to the user, we substitute indices by their related names.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Syntax}
\label{syntax}
We let $x, y,\ldots$ range over \emph{channel names} or \emph{variables} and
$P, Q,\ldots$ over processes.
The syntax of the \picalc{} \cite{Sangio01} is given by the following grammar.
\begin{equation*}
  \begin{split}
    P,Q ::=      & \; \PO       &\text{(terminated process)}  \\
                |& \; (\new{}x)  P   & \text{(restriction)} \\
                |& \; \comp{P}{Q}  & \text{(parallel composition)} \\
                |& \; \recvp{x}{y}.P  & \text{(input)} \\
                |& \; \sendp{x}{y}.P   &\text{(output)}
  \end{split}
\end{equation*}
Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ is the restriction process where channel $x$ is created new and is bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}.P$ and $\sendp{x}{y}.P$ denote respectively the input and output processes of a variable/channel $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x)P$ and input $\recvp{x}{y}.P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need a way to deal with bound names in these continuation processes.
Channel names are cumbersome to mechanise: inserting a new variable into a context means proving that the name of such variable is different to all other variable names in context.
Instead we adopt de Bruijn indices \cite{} where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\todo{rewrite the sytanx below, following the syntax above}
\begin{example}
\hfill{}\\
The \picalc{} term \hfill{} $(\textbf{new} \; x) (x ( y ) . y \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new} \; y) (x \langle y \rangle . y ( z ) .\mathbf{0}))$ \\ is represented as \hfill{} $(\textbf{new}) (0 () . 0 \langle \texttt{"Hello world!"} \rangle . \mathbf{0} \parallel (\textbf{new}) (1 \langle 0 \rangle . 0 () . \mathbf{0}))$.
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

A variable occurring under $n$ abstractions has $n$ possible variables to refer to.
References outside of that range are meaningless.
It is useful to rule out these nonsensical terms syntactically.
\todo{replace ref with autoref everywhere...}
In  \autoref{fig:var} we do so by introducing the indexed family of types $\Var_n$: for all naturals $n$, the type $\Var_n$ has $n$ distinct elements.

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
Figure \ref{fig:syntax} shows the syntax of \picalc{} in Agda. We increase the \emph{depth} counter by 1 in channel restriction and input processes.

@@

\begin{equation*}
\begin{split}
  \Process_n ::=& \; \PO_n                    \\
              |& \; \new{} \Process_{\suc n}          \\
              |& \; \comp{\Process_n}{\Process_n}          \\
              |& \; \recv{\Var_n}{}\Process_{\suc n} \\
              |& \; \send{\Var_n}{\Var_n}\Process_n
\end{split}
\end{equation*}
\caption{Well-scoped grammar for \picalc{} using de Bruijn indices}
\label{fig:syntax}
\end{figure}

Using type-level de Bruijn indices makes our syntax well-scoped by construction --- the semantics of our language can now be defined on the totality of the syntax.
In order to make the de Bruijn indices representation more human-readable, we can annotate these indices with the original channel names.
Machines can then manipulate references mechanically whilst using channel names to present \picalc{} terms back to humans.
